The Evolution of Phenotypically Invariant Genetic Networks

### 2021-05-06 kalman_walk_paper.tex
appendix

Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
\begin{equation}
    \begin{aligned}
        D(A, B, C)^2
        &:= 
        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
        \right\} dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
        \right\} dt  ,
        % &= 
        % \int_0^\infty \rho(t)  \tr\left\{
        %         C e^{At} B B^T e^{A^T t} C^T 
        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
        %         \right. \\ &\qquad \qquad \left. {}
        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
        % \right\} dt \\
        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
        % &= 
        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
    \end{aligned}
\end{equation}
where $\tr X$ denotes the trace of a square matrix $X$.
How does this change as we perturb about $(A_0, B_0, C_0)$?
First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
Since
\begin{equation}
  \begin{aligned}
      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
      &=
      \int_0^t e^{As} Z e^{A(t-s)} ds, 
  \end{aligned}
\end{equation}
we have that
\begin{equation}
  \begin{aligned}
      \frac{d}{du} D(A+uZ,B_0,C_0)^2 \vert_{u=0}
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
  \end{aligned}
\end{equation}
and, by differentiating this and supposing that $A$ is on the optimal set,
i.e., $h_A(t)=h_0(t)$, (so without loss of generality, $A=A_0$):
\begin{equation}
  \begin{aligned}
      \calH^{A,A}(Y,Z) 
      &:= 
      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)^2 \vert_{u=v=0} \\
      &=
        \int_0^\infty \rho(t) \tr\left\{ C_0 
        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
        B_0 B_0^T 
        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
        C_0^T \right\} dt  .
  \end{aligned}
\end{equation}

>>>>

The fitness of a system 
depends on the difference between the system's impulse response and the optimal impulse response,
measured  as a weighted sum of
the distance between the impulse response from optimal.
With $\rho(t) \ge 0$ a weighting function on $[0,\infty)$,
and $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set,
the distance is equation \eqref{eqn:distance}:
\begin{align}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
To see how this is affected by small changes to the system,
first note that if $\|\cdot\|_\rho$ is the $L^2(\rho)$ norm,
\begin{align}
	\|g\|_\rho^2 := \int_0^\infty \rho(t) \|g(t)\|^2 dt
\end{align}
then the distance to optimum of a perturbed system, $h(t) = h_0(t) + \epsilon g(t)$, is
\begin{align}
	D(h_0 + \epsilon g) = \epsilon \|g\|_\rho .
\end{align}

Now consider the perturbed systems $(A(\epsilon), B(\epsilon), C(\epsilon)) = (A(0), B(0), C(0)) + \epsilon (U, V, W)$,
and let $\h_\epsilon(t) = C(\epsilon) e^{A(\epsilon) t} B(\epsilon)$.
If two populations with systems $(A(0), B(0), C(0))$ and $(A(\epsilon), B(\epsilon), C(\epsilon))$ interbreed,
then the mean resulting system is $(A({\epsilon/2}), B({\epsilon/2}), C({\epsilon/2}))$,
but with greater variance in the $F_2$ than the $F_1$ offspring (see Figure \ref{fig:conceptual_fig}).
Indeed, if the difference between (say) $A(0)_{ij}$ and $A(\epsilon)_{ij}$
is due to $n$ fixed differences with additive effect size $\epsilon/(2n)$ each,
then the standard deviation of $A_{ij}$ among the $F_1$ offspring is equal to that
within each parent population, while among the $F_2$ offspring it is $\epsilon / \sqrt{8 n}$.
This implies that the impulse response, $h(t)$, of an $F_2$ will also differ from optimum
by something of order $\epsilon$.
However, since the displacement between the populations lies along a ridge,
the $F_1$ offspring are closer:
the Taylor expansions for $h$ about both 0 and $\epsilon$ are
\begin{align}
    h_{\epsilon/2}(t)
    &= h_0(t)
        + \frac{\epsilon}{2} \partial_\epsilon h_0(t)
        + \frac{\epsilon^2}{8} \partial_\epsilon^2 h_0(t)
        + O(\epsilon^3) \\
    \text{and} \quad
    &= h_\epsilon(t)
        - \frac{\epsilon}{2} \partial_\epsilon h_\epsilon(t)
        + \frac{\epsilon^2}{8} \partial_\epsilon^2 h_\epsilon(t)
        + O(\epsilon^3)
\end{align}
Since $h_\epsilon(t) = h_0(t)$ by assumption,
and $\partial_\epsilon h_\epsilon(t) = \partial_\epsilon h_0(t) + \epsilon \partial_\epsilon^2 h_0(t) + O(\epsilon^2)$,
combining these expressions we get that
\begin{align}
    h_{\epsilon/2}(t)
    &= h_0(t)
        - \frac{\epsilon^2}{8} \partial_\epsilon^2 h_0(t)
        + O(\epsilon^3) .
\end{align}
In other words, if the parental populations are both optimal but differ by $\epsilon$,
then the $F_1$ offspring differ from optimal by something of order $\epsilon^2$,
but the $F_2$ by something of order $\epsilon$.

Of course, when moving away from the ridge, the fitness landscape is flatter in some directions than others,
and how much so depends on the details.
For completeness, we describe here how to compute the sensitivity of distance to optimum
with respect to each of $A$, $B$, and $C$.
Changing notation slightly, let's rewrite the squared distance to optimum
of a system $(A, B, C)$:
\begin{equation}
    \begin{aligned}
        D(A, B, C)^2
        &:= 
        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
        \right\} dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
        \right\} dt  ,
        % &= 
        % \int_0^\infty \rho(t)  \tr\left\{
        %         C e^{At} B B^T e^{A^T t} C^T 
        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
        %         \right. \\ &\qquad \qquad \left. {}
        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
        % \right\} dt \\
        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
        % &= 
        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
    \end{aligned}
\end{equation}
where $\tr X$ denotes the trace of a square matrix $X$.
How does this change as we perturb about $(A_0, B_0, C_0)$?
First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
Since
\begin{equation}
  \begin{aligned}
      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
      &=
      \int_0^t e^{As} Z e^{A(t-s)} ds, 
  \end{aligned}
\end{equation}
we have that
\begin{equation}
  \begin{aligned}
      \frac{d}{du} D(A+uZ,B_0,C_0)^2 \vert_{u=0}
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
  \end{aligned}
\end{equation}
and, by differentiating this and supposing that $A$ is on the optimal set,
i.e., $h_A(t)=h_0(t)$, (so without loss of generality, $A=A_0$):
\begin{equation}
  \begin{aligned}
      \calH^{A,A}(Y,Z) 
      &:= 
      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)^2 \vert_{u=v=0} \\
      &=
        \int_0^\infty \rho(t) \tr\left\{ C_0 
        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
        B_0 B_0^T 
        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
        C_0^T \right\} dt  .
  \end{aligned}
\end{equation}
